Formal Verification of Real-Time Data Processing of the LHC Beam Loss Monitoring System: A Case Study

### 2011-03-29 fmics/paper.tex
merge changes from dropbox

%Figure~\ref{fig:theorems} gives a selection of theorems we have proved in HOL, some of which refer to additional constants defined in Figure~\ref{fig:auxdefs}.
Figure~\ref{fig:auxdefs} defines a set of additional functions required to prove our main results.
The function $\mathsf{delay\_sum}\;n$ represents the cumulative delay of the preceding shift registers of a slice.
For example, $\mathsf{delay\_sum}\;4 = \mathsf{delay}\;3 + \mathsf{delay}\;1 + \mathsf{delay}\;0 = 2 +1 +1 = 4$.
The function $\mathsf{last\_update}\;n\;t$ specifies the time that the $n$-th slice last updated, and $\mathsf{exact}\;D\;n\;x\;t$ computes the exact sum of consecutive input counts, without delay, that $\mathsf{output}\;D\;n\;x\;t$ is supposed to approximate (so $\mathsf{exact}$ is to $\mathsf{true\_sum}$ as $\mathsf{output}$ is to $\mathsf{RS}$).

%Our central result, Theorem~(\ref{output_input_at_update_times}), equates the output of a slice to a sum of consecutive input counts.
Our central result equates the output of a slice to a sum of consecutive input counts.
More precisely, it says that if slice $n$ was just updated, then the output at tap $x$ is equal to the sum of $((\mathsf{tap}\;n\;x)+1)\times(\mathsf{delay}\;n)$ input values, starting from the input $\mathsf{delay\_sum}\;n$ time steps ago.
%Note that the precise statement of such a theorem is complicated by details of the value of the $\mathsf{output}$ at times when the shift registers have not yet received any values.

%To prove Theorem~(\ref{output_input_at_update_times}), we first make the following observations\footnote{The precise statements of these observations are complicated by details of what happens when $t$ is low, or equivalently, when we are looking at shift register cells that have not yet received any input.}

\begin{theorem}\label{output_input_at_update_times}
For all values of $D$, $n$, and $x$, the $\mathsf{output}$ of slice $n>0$ at an update time $t$ satisfies
\begin{equation*}%\label{output_input_at_update_times}
%\forall{D\,n\,x\,t}.\,{0 < n}\land{\mathsf{update\_time\;n\;t}}\implies\\
\mathsf{output}\;D\;n\;x\;t=\sum_{m=0}^{((\mathsf{tap}\;n\;x)+1)\times(\mathsf{delay}\;n)-1}\begin{cases}0 \qquad\qquad\qquad  t<m+\mathsf{delay\_sum}\;n\\D\;(t-m-\mathsf{delay\_sum}\;n)\quad\text{otherwise}\end{cases}
\end{equation*}
\end{theorem}

\begin{proof}
The shift register cells get their values from the previous cell but they only update every $\mathsf{delay}\;n$ time steps. Therefore\footnote{We do not go through the details here of what happens when $t$ is small, for example $t<m\times\mathsf{delay}\;n$.},
\begin{equation*}
\mathsf{SR}\;D\;n\;(m+1)\;t=\mathsf{SR}\;D\;n\;m\;(t-\mathsf{delay}\;n)
\end{equation*}
By induction on $m$, using the above,
\begin{equation*}
\mathsf{SR}\;D\;n\;m\;t=\mathsf{SR}\;D\;n\;0\;(t-(m\times\mathsf{delay}\;n))
\end{equation*}
By induction on $t$, we can show that $\mathsf{output}\;D\;n\;x\;t$ is a sum of values of consecutive shift register cells
\begin{equation*}
\mathsf{output}\;D\;n\;x\;t=\sum_{m=0}^{\mathsf{tap}\;n\;x}\mathsf{SR}\;D\;n\;m\;t
\end{equation*}
Combining the last two results, $\mathsf{output}\;D\;n\;x\;t$ is a sum of consecutive values of the first shift register cell, $\mathsf{SR}\;D\;n\;0$.
Thus, we can express $\mathsf{output}\;D\;n\;x\;t$ in terms of $\mathsf{output}\;D\;n'\;x'\;t$, where $(n',x')=\mathsf{input}\;n$, since the source of $\mathsf{SR}\;D\;n\;0$ is the previous (input) slice's output.
Finally, by induction on $n$, we can express $\mathsf{output}\;D\;n\;x\;t$ as a function of $D$ alone, as required, using the result above for the inductive case.
\qed
\end{proof}

Since a slice's output stays constant between update times for the slice, Theorem~\ref{output_input_at_update_times} characterises all outputs at all times.
Thus, the SRS structure's outputs are exact running sums of the input counts over windows whose sizes depend on the the width of the slice and position of taps.
However, the sums are delayed by the total delay across all previous slices (represented by $\mathsf{delay\_sum}$).

@@

\begin{theorem}\label{output_eq_exact}
For all values of $D$, $n$, and $x$, the $\mathsf{output}$ of slice $n>0$ at all times $t$ satisfies
\begin{equation*}
\mathsf{output}\;D\;n\;x\;t=\begin{cases}0\qquad\qquad\qquad\qquad\qquad\quad\mathsf{last\_update}\;n\;t+1<\mathsf{delay\_sum}\;n\\\mathsf{exact}\;D\;n\;x\;(\mathsf{last\_update}\;n\;t+1-\mathsf{delay\_sum}\;n)\quad\text{otherwise}\end{cases}
\end{equation*}
\end{theorem}
\begin{proof}
Follows from Theorem~\ref{output_input_at_update_times}, using the definition of $\mathsf{exact}$, and the fact that $\mathsf{output}\;D\;n\;x\;t$ has the same value as at the last update time.\qed
\end{proof}

@@

%\begin{figure}
%\caption{
%A selection of theorems about our model proved using HOL4.
%\label{fig:theorems}
%}
%\begin{align}
%\label{update_time_last_update}
%&\forall{n\,t}.\,\mathsf{update\_time}\;n\;(\mathsf{last\_update}\;n\;t)\\
%%\label{last_update_thm}
%%&\forall{n\,t}.\,\mathsf{last\_update}\;n\;t=t-t\operatorname{mod}\mathsf{delay}\;n\\
%%SR_prev
%%SR_first
%\label{output_last_update}
%&\forall{D\,n\,x\,t}.\,\mathsf{output}\;D\;n\;x\;t=\mathsf{output}\;D\;n\;x\;(\mathsf{last\_update}\;n\;t)\\
%\label{output_sum}
%&\forall{D\,n\,x\,t}.\,{0<n}\implies\mathsf{output}\;D\;n\;x\;t=\sum_{m=0}^{\mathsf{tap}\;n\;x}\mathsf{SR}\;D\;n\;m\;t
%\end{align}
%%\begin{multline}
%%\label{output_source_at_update_times}
%%\forall{D\,n\,x\,t}.\,{0<n}\land{\mathsf{update\_time\;n\;t}}\implies\\
%%\mathsf{output}\;D\;n\;x\;t=\sum_{m=0}^{\mathsf{tap}\;n\;x}\begin{cases}0&t\leq m\times(\mathsf{delay}\;n)\\\mathsf{source}\;D\;n\;0\;(t-m\times(\mathsf{delay\;n})-1)&\text{otherwise}\end{cases}
%%\end{multline}
%\begin{multline}
%\label{output_prev_at_update_times}
%\forall{D\,n\,x\,t}.\,{0<n}\land{\mathsf{update\_time}\;n\;t}\implies\\
%\mathsf{output}\;D\;n\;x\;t=\sum_{m=0}^{\mathsf{tap}\;n\;x}\begin{cases}0&t < m\times\mathsf{delay}\;n+\mathsf{delay}\;n'\\\mathsf{output}\;D\;n'\;x'\;(t-m\times\mathsf{delay}\;n-\mathsf{delay}\;n')&\text{otherwise}\end{cases}\\
%\qquad\text{where $(n',x')=\mathsf{input}\;n$}
%\end{multline}
%\begin{multline}\label{output_eq_exact}
%\forall{D\,n\,x\,t}.\,{0 < n}\implies\\\mathsf{output}\;D\;n\;x\;t=\begin{cases}0&\mathsf{last\_update}\;n\;t<\mathsf{delay}\;n\\\mathsf{exact}\;D\;n\;x\;(\mathsf{last\_update}\;n\;t+1-\mathsf{delay\_sum}\;n)\end{cases}
%\end{multline}
%%max_error
%\end{figure}

>>>>

Figure~\ref{fig:auxdefs} defines a set of additional functions required to prove our main results.
The function $\mathsf{delay\_sum}\;n$ represents the cumulative delay of the preceding shift registers of a slice.
For example, $\mathsf{delay\_sum}\;4 = \mathsf{delay}\;3 + \mathsf{delay}\;1 + \mathsf{delay}\;0 = 2 +1 +1 = 4$.
The function $\mathsf{last\_update}\;n\;t$ specifies the time that the $n$-th slice last updated, and $\mathsf{exact}\;D\;n\;x\;t$ computes the exact sum of consecutive input counts, without delay, that $\mathsf{output}\;D\;n\;x\;t$ is supposed to approximate.
% (so $\mathsf{exact}$ is to $\mathsf{true\_sum}$ as $\mathsf{output}$ is to $\mathsf{RS}$).

Our central result equates the output of a slice to a sum of consecutive input counts.
More precisely, it says that if slice $n$ was just updated, then the output at tap $x$ is equal to the sum of $((\mathsf{tap}\;n\;x)+1)\times(\mathsf{delay}\;n)$ input values, starting from the input $\mathsf{delay\_sum}\;n$ time steps ago.

\begin{theorem}\label{output_input_at_update_times}
For all values of $D$, $n$, and $x$, the $\mathsf{output}$ of slice $n>0$ at an update time $t$ satisfies
\begin{equation*}
\mathsf{output}\;D\;n\;x\;t=\sum_{m=0}^{((\mathsf{tap}\;n\;x)+1)\times(\mathsf{delay}\;n)-1}\begin{cases}0 \qquad\qquad\qquad  t<m+\mathsf{delay\_sum}\;n\\D\;(t-m-\mathsf{delay\_sum}\;n)\quad\text{otherwise}\end{cases}
\end{equation*}
\end{theorem}

\begin{proof}
The shift register of slice $n$ is updated every $\mathsf{delay}\;n$ time steps.
When updated, the values in its cells are shifted one cell.
%The shift register cells get their values from the previous cell but they only update every $\mathsf{delay}\;n$ time steps. Therefore
Therefore\footnote{For presentation convenience, here, we do not provide the details regarding the case when $t$ is small, for example $t<m\times\mathsf{delay}\;n$.
For the full description of the theorems and proof script, please refer to {\bf Can we put a link pointing to the proof script here?}},
\begin{equation*}
\mathsf{SR}\;D\;n\;(m+1)\;t=\mathsf{SR}\;D\;n\;m\;(t-\mathsf{delay}\;n)
\end{equation*}
By induction on $m$, using the above,
\begin{equation*}
\mathsf{SR}\;D\;n\;m\;t=\mathsf{SR}\;D\;n\;0\;(t-(m\times\mathsf{delay}\;n))
\end{equation*}
By induction on $t$, we can show that $\mathsf{output}\;D\;n\;x\;t$ is a sum of values of consecutive shift register cells
\begin{equation*}
\mathsf{output}\;D\;n\;x\;t=\sum_{m=0}^{\mathsf{tap}\;n\;x}\mathsf{SR}\;D\;n\;m\;t
\end{equation*}
Combining the last two results, $\mathsf{output}\;D\;n\;x\;t$ is a sum of consecutive values of the first shift register cell, $\mathsf{SR}\;D\;n\;0$.
Thus, we can express $\mathsf{output}\;D\;n\;x\;t$ in terms of $\mathsf{output}\;D\;n'\;x'\;t$, where $(n',x')=\mathsf{input}\;n$, since the source of $\mathsf{SR}\;D\;n\;0$ is the previous (input) slice's output.
Finally, by induction on $n$, we can express $\mathsf{output}\;D\;n\;x\;t$ as a function of $D$ alone, as required, using the result above for the inductive case.
\qed
\end{proof}

Since the outputs of a slice stay constant between update times for the slice, Theorem~\ref{output_input_at_update_times} suffices to characterize all outputs at all times.
Thus, the SRS structure's outputs are equal to the exact running sums of the input counts over windows whose sizes depend on the the width of the slice's shift register and position of the taps.
However, the sums are delayed by the total delay across all previous slices (represented by $\mathsf{delay\_sum}$).

@@

\begin{theorem}\label{output_eq_exact}
For all values of $D$, $n$, and $x$, the $\mathsf{output}$ of slice $n>0$ at all times $t$ satisfies
\begin{equation*}
\mathsf{output}\;D\;n\;x\;t=\begin{cases}0\qquad\qquad\qquad\qquad\qquad\quad\mathsf{last\_update}\;n\;t+1<\mathsf{delay\_sum}\;n\\\mathsf{exact}\;D\;n\;x\;(\mathsf{last\_update}\;n\;t+1-\mathsf{delay\_sum}\;n)\quad\text{otherwise}\end{cases}
\end{equation*}
\end{theorem}
\begin{proof}
The proof follows from Theorem~\ref{output_input_at_update_times}, using the definition of $\mathsf{exact}$, and the fact that $\mathsf{output}\;D\;n\;x\;t$ has the same value as at the last update time.\qed
\end{proof}

Note that the definition of the $\mathsf{true\_sum}$ is easily defined in terms of $\mathsf{exact}$, that is $\mathsf{true\_sum}\;D\;n\;t = \mathsf{exact}\;D\;\left(\left\lfloor\frac{n}{2}\right\rfloor+1\right)\;(n\operatorname{mod}2)\;t$.
Sine the $\mathsf{RS}$ is also similarly defined in terms of the $\mathsf{output}$, Theorem~\ref{output_eq_exact} establishes a relationship between the values of the $\mathsf{RS}$ and $\mathsf{true\_sum}$.

We formulate the difference between the $\mathsf{RS}$ values and their corresponding $\mathsf{true\_sum}$s, caused by the delay, as $\mathsf{error}$. 
In order to define a maximum bound on the $\mathsf{error}$, we had to formulate some characterization of the input stream to the SRS structure as constraints on $D$.
One of these constraints is the bound on the difference between the two consecutive count values.

@@

%\begin{figure}
%\caption{
%A selection of theorems about our model proved using HOL4.
%\label{fig:theorems}
%}
%\begin{align}
%\label{update_time_last_update}
%&\forall{n\,t}.\,\mathsf{update\_time}\;n\;(\mathsf{last\_update}\;n\;t)\\
%%\label{last_update_thm}
%%&\forall{n\,t}.\,\mathsf{last\_update}\;n\;t=t-t\operatorname{mod}\mathsf{delay}\;n\\
%%SR_prev
%%SR_first
%\label{output_last_update}
%&\forall{D\,n\,x\,t}.\,\mathsf{output}\;D\;n\;x\;t=\mathsf{output}\;D\;n\;x\;(\mathsf{last\_update}\;n\;t)\\
%\label{output_sum}
%&\forall{D\,n\,x\,t}.\,{0<n}\implies\mathsf{output}\;D\;n\;x\;t=\sum_{m=0}^{\mathsf{tap}\;n\;x}\mathsf{SR}\;D\;n\;m\;t
%\end{align}
%%\begin{multline}
%%\label{output_source_at_update_times}
%%\forall{D\,n\,x\,t}.\,{0<n}\land{\mathsf{update\_time\;n\;t}}\implies\\
%%\mathsf{output}\;D\;n\;x\;t=\sum_{m=0}^{\mathsf{tap}\;n\;x}\begin{cases}0&t\leq m\times(\mathsf{delay}\;n)\\\mathsf{source}\;D\;n\;0\;(t-m\times(\mathsf{delay\;n})-1)&\text{otherwise}\end{cases}
%%\end{multline}
%\begin{multline}
%\label{output_prev_at_update_times}
%\forall{D\,n\,x\,t}.\,{0<n}\land{\mathsf{update\_time}\;n\;t}\implies\\
%\mathsf{output}\;D\;n\;x\;t=\sum_{m=0}^{\mathsf{tap}\;n\;x}\begin{cases}0&t < m\times\mathsf{delay}\;n+\mathsf{delay}\;n'\\\mathsf{output}\;D\;n'\;x'\;(t-m\times\mathsf{delay}\;n-\mathsf{delay}\;n')&\text{otherwise}\end{cases}\\
%\qquad\text{where $(n',x')=\mathsf{input}\;n$}
%\end{multline}
%\begin{multline}\label{output_input_at_update_times}
%\forall{D\,n\,x\,t}.\,{0 < n}\land{\mathsf{update\_time\;n\;t}}\implies\\\mathsf{output}\;D\;n\;x\;t=\sum_{m=0}^{((\mathsf{tap}\;n\;x)+1)\times(\mathsf{delay}\;n)-1}\begin{cases}0&t<m+\mathsf{delay\_sum}\;n\\D\;(t-m-\mathsf{delay\_sum}\;n)&\text{otherwise}\end{cases}
%\end{multline}
%\begin{multline}\label{output_eq_exact}
%\forall{D\,n\,x\,t}.\,{0 < n}\implies\\\mathsf{output}\;D\;n\;x\;t=\begin{cases}0&\mathsf{last\_update}\;n\;t<\mathsf{delay}\;n\\\mathsf{exact}\;D\;n\;x\;(\mathsf{last\_update}\;n\;t+1-\mathsf{delay\_sum}\;n)\end{cases}
%\end{multline}
%%max_error
%\end{figure}
